Replacing handwritten signatures with open electronic signature software

### 2020-05-07 paper/main.tex
Add final draft for the SVK

\documentclass[thesismargins, english, thesislinespacing, twoside, openright, upjsfrontpage]{rnthesis}
\usepackage[english]{babel}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}

@@

\title{Replacing handwritten signatures with open electronic signature software}
\author{Jakub Ďuraš}
\typprace{bachelor's}
\rok{2020}
\odbor{Applied Informatics}
\miesto{Košice}
\podakovanie{
  Thanks to the Planet!
}
\veduci{RNDr. Viliam Kačala}
\pracovisko{Institute of Computer Science}
\pdfzadanie{zadanie.pdf}

\abstract{
With the recent changes in the legal status of electronic signatures in many parts of the world, there is a need for easily accessible solutions intended as an alternative to handwritten signatures. This may be more necessary than ever since electronic communication is the preferred way of communication. This bachelor thesis aims to explore the principles and legal status of electronic signatures, review current electronic signature software, and explore possible obstacles the open-source community is facing to develop such specialized applications. We propose an open-source, cross-platform, and user-friendly software compliant with the eIDAS Regulation (Regulation No 910/2014). Our application should allow ordinary users to quickly sign and verify signatures of different types of documents and therefore easily use them in everyday life.
}

@@

\abstrakt{
Abstrakt v SK jazyku bude pridaný neskôr ako voľný preklad z EN verzie (pozri ďalšiu stranu). Predíde sa tak zbytočnej práci keďže EN abstrakt sa môže časom meniť.
}

@@

\begin{document}
\maketitle
\newpage
\tableofcontents
\listoffigures
\listoftables
% zoznam značiek a skratiek. Pre tento koncept neexistuje
% LaTeXovsky príkaz

@@

Will be added at the end. Lorem ipsum dolor sit amet, consectetuer adipiscing elit.
Integer lacinia, nulla porta varius tempus, lacus metus blandit
lorem, a rutrum justo wisi id sapien. Integer risus libero,
feugiat eleifend, ornare ac, volutpat nec, sem. In facilisis,
quam eu elementum aliquet, lorem quam euismod dui, aliquet
laoreet purus ipsum ac quam.

\chapter{Background and theory}

In our work, we rely on knowledge in several different subjects that we go over in the following sections.
TODO: Sounds very meh for the very first sentece of the thesis.
Firstly, the current state of the law and recent changes relating to the use of electronic signatures.
Secondly, on the well-established concepts in cryptography like asymmetric cryptography or hashing.
Lastly, expertise from the field of software engineering - especially design, implementation, and maintenance of computer software.
All in relation to the electronic signatures, and, where applicable, open-source and desktop computer software.

TODO: Think about adding section on handwritten signatures

@@

Unless specified otherwise, we are considering law applicable locally in the Slovak Republic (SR).
Its law is greatly influenced by the European Union (EU), being its member since 2004, and to some extent by the rest of the world.
Therefore, we can assume this is at least partially applicable outside of the SR.

@@

Signatures are essential part of the written legally binding documents like contracts.
They are permanently affixed to the document and are supposed to uniquely identify the person and its deliberate, informed consent.
As can be seen in the Slovak Civil Code, "A written legal act is valid if signed by the acting person;" \cite{civilcode}.
Law often explicitly requires signatures and further clarifies their expected use.
For example, when selling an enterprise, "The contract requires written form and attested signatures of the seller and the buyer." \cite{civilcode}.

Signature is considered an "attested signature"\footnote{Translation from "osvedčený podpis" as used in the Slovak law.} if it is attested by the authorized third party.
Such process, known as legalization, is regulated by the law and its purpose is to attest information which could form the basis for the exercise of rights or which could cause legal consequences \cite{notarylaw}.

With regards to electronic signatures, law within the EU used to differ, with law applicable in the SR being now repealed "Act No. 215/2002 Coll.".

@@

Advanced electronic signatures are a subset of electronic signatures that have to \textbf{uniquely link and identify} the signature author, be created in a way that is \textbf{possible only by them} and any \textbf{changes to the signed document have to be detectable} \cite{eidas}.
From the technical point of view, standards PAdES, XAdES, and CAdES specified by the European Telecommunications Standards Institute (ETSI) comply with these requirements. We explain why this is the case in Chapter \ref{cryptography}.

Qualified electronic signatures (QES) are a subset of advanced electronic signatures with more specific requirements that can be found in the annex 1 of the eIDAS regulation\footnote{Available at https://eur-lex.europa.eu/legal-content/EN/TXT/HTML/?uri=CELEX:32014R0910\#d1e32-111-1.}.
In more practical terms, such advanced electronic signatures are created with a qualified device using a qualified certificate and a qualified trust service. Qualified in this case means authorized for such use by the legal authorities.

Furthemore, a qualified electronic signature has the legal effect of a handwritten signature \cite{eidas} and attested handwritten signatures in the SR.

Mentioned terminology is used when considering natural person. Legal entities are able to use "electronic seals", which are, from the technical point of view and our purposes, identical.

Overall, electronic signatures are supposed to be, within the EU, interoperable and transparent alternative to hand-written signatures.

\subsection{Software copyright}

@@

Such right can be exercised, as far as we are concerned, to license the software under either proprietary or open-source license.
Proprietary meaning under exclusive legal right of the author, typically also confidential and distributed as a paid product.
Open-source meaning having its source code freely available, typically also distributed for free and without any liability (the software is provided "as is").

Motivation to license the software under an open-source license can differ, and so do such licenses.
While a proprietary license is usually made specifically for that entity and its interests, open-source licenses tend to be reused between different authors.
This means consumers of the software can quickly recognize their rights and responsibilities if they decide to use, modify, or distribute the software.
In general, we can classify the open-source licenses into two categories: permissive and copyleft.

Copyleft license, in general, requires the user to publish the modified work under the compatible (free) license.
Meaning forcing them to extend the rights they have received onto others and, in turn, somehow hindering use in software licensed under a different license.
Popular example of such license is the GNU General Public License (GPL) and its derivatives like GNU Lesser GPL (LGPL) or GNU Affero GPL (AGPL).

Permissive license, on the other hand, does not have such requirement in place and is more suitable for potential commercialization.
They typically allow commercial use, modifications, distribution or sublicensing as long as the author is not held liable and the original copyright and license is distributed with the software.
Often used permissive licenses are MIT License, Apache License or BSD License.

It is not uncommon to see hybrid licensing - licensing under more than one license, with one being typically open-source and one proprietary.
In this scenario, users can choose which license they want to use based on their needs.

@@

As we have already mentioned, software licenses can influence whether and how it can be used as a part of different software.
Both when we are using someone else's software and when someone else is using our software.
Generally speaking, the more open the license is, the higher the probability it will not stop adoption for legal reasons.
Various possible combinations can be seen in Figure \ref{fig:license-compatibility}, where greyed out options mean such combination is not possible unless a separate licensing agreement is reached with the copyright owner.

\obrazok{./figures/license-compatibility}{Schematic representation of license directionality.\cite{licensing}}{license-compatibility}

@@

Previously mentioned standards PAdES, XAdES, and CAdES define how asymmetric cryptography, hashing, or timestamping should be used to comply with the standards.
We go over a few of these topics and why they are able to satisfy the requirements.

@@

There are several issues that make symmetric encryption unsuitable. Let us imagine we are creating a key for each pair of users in our network - imaginary country Isle of Alice and Bob - for symmetric encryption and decryption of documents shared between each Alice and Bob.

\textbf{Secure distribution of keys} seems to be feasible when we are adding a new user to our network. We require each user to personally come and pick up physical device that contains pregenerated keys. Changes made after that, however, would mean we either need to require periodic personal visit to update the saved keys or transmission of these keys over some medium which can be potentialy unsafe. Any new members of the network are unable to participate untill other Alices and Bobs get their updates.

@@

which means that even our little country with a population of only 70 000 requires roughly 2.5 billion keys. However, since our signature system needs to be portable, we will need to count also with a population of other EU countries, and the number of keys quickly explodes.

\begin{figure}[!htb]
  \centering

@@

  \centerline{\parbox{13cm}{\sl \caption{Cubic growth of the number of keys.}}}
\end{figure}

@@

These issues can be addressed with asymmetric cryptography.
In general, we assign a pair of keys to each person, one of them used for encryption (public key), another used for decryption (private key).
One party can encrypt, and the other is the only one that can decrypt.

Digital signatures are a specific public-key algorithms.
The signer is the only one who poses private key used for signing, and others can use the public key for verifying of signatures as can be seen in Figure \ref{fig:pubkey-scheme} where \(x\) is the "document", and \(s\) is the signature that is sent with the document.
Alice can then sign the document \textit{including} Bob's signature.

\obrazok{./figures/digital-signature-scheme}{Principle of digital signatures which involves signing and verifying a message (Paar 2009). TODO: I don't have rights to this image!!!}{pubkey-scheme}

Practical, widely used implementation of this idea is the RSA signature scheme that relies on the integer factorization problem (difficulty of factoring a product of two large primes) published in 1978 \cite{rsa}.

In Slovakia, the private key is usually handed over to the user personally, saved on the ID card. This ID card (used together with a card reader) provides an API for safe access, management, and cryptoprocessing. Such device is in general called a Hardware Security Module (HSM), and the API relevant for this thesis is PKCS \#11 \footnote{PKCS \#11 specification available at \url{http://docs.oasis-open.org/pkcs11/pkcs11-base/v2.40/os/pkcs11-base-v2.40-os.html}.}.

We are left with the problem of public key distribution - even though public-key schemes do not require a \textit{secure channel}, they require \textit{authenticated channels} for the distribution of the public keys \cite{cryptotxtbook}.

Common solution is the use of \textit{certificates}. Certificates are essentially digital signatures with metadata that can be used to establish validity of the received public key before it is used.
Distribution of such certificates is the responsibility of a Certification Authority (CA) - a third party that all users in the network trust. In the case of the Slovak ID cards, it is Disig - SVK eID ACA issued by the National Security Authority (NBÚ in Slovak).
Such certificates therefore reliably uniquely link and identify the public key owners as can be seen in Figure \ref{fig:certificate-scheme}\footnote{Created by unknown author, licensed under the CC BY-SA 3.0 license.}.

\obrazok{./figures/certificate-scheme}{Application of digital signature with certificate. TODO: Modify to add CA, remove hashing}{certificate-scheme}

Commonly used standard for public key certificates is X.509\footnote{Specification available under RFC5280 at \url{https://tools.ietf.org/html/rfc5280}.}.

@@

It is often useful to map data of arbitrary size to fixed-length (typically short) set of bits. This process is commonly called \textit{hashing}, function that can be used for it \textit{hash function}, and output of such function \textit{hash}.

In cryptography, we are considering only hash functions that can also satisfy these additional requirements:

\begin{enumerate}
  \item Deterministic - same input data are always mapped to the same hash.
  \item One-way - it should be practically impossible to retrieve input data for a given hash.
  \item Resistant to collisions - it should be practically impossible to find two inputs with the same hash.
  \item Even small change to the input should lead to significantly different hash - so-called \textit{avalanche effect}.
\end{enumerate}

In the context of digital documents, we can use such hash function to generate a hash for the document.
This hash can represent a document of any size and is used during the creation of a digital signature instad of document itself.
Any further changes to the document will, therefore, lead to a significantly different hash, which means the digital signature will no longer be valid, and these changes are detected.
For this purpose, only hash functions that are reasonably fast can be used, so that we can utilize them on documents of variable size.

TODO: Add scheme of document > hash - valid signature, change in document > different hash - invalid signature.

An example of commonly used hash function family is Secure Hash Algorithms (SHA), with hash functions relevant to this thesis being mainly SHA-2\footnote{SHA-2 specification available in FIPS PUB 180-4 at \url{https://nvlpubs.nist.gov/nistpubs/FIPS/NIST.FIPS.180-4.pdf}.}.

@@

A common part of the paper signed documents is the date when they were signed.
Their digital counterpart is PKI-based \textit{trusted timestamping} - a process of digitally signing date and time of document creation or modification.
No one should be able to tamper with such timestamp - not even the author of the document.
Additionally, we should not be able to choose arbitrary date and time ourselves, but it should represent the actual time at the signing.

To achieve this, a trusted third party - Timestamping Authority (TSA) - is the one that will sign the hash of the document together with the current date and time.
This process can be seen in Figure \ref{fig:trusted-timestamping}\footnote{Schema created by Bart Van den Bosch, licensed under the CC BY-SA 2.0 BE license.}.
Verifying can be by decrypting this information using the public key and comparing the hashes. Any attempt at tampering is therefore detected.

@@

The commonly used standard for the PKI-based timestamping is part of the X.509\footnote{Specification available under RFC3161 at \url{https://www.ietf.org/rfc/rfc3161.txt}.}.

@@

Extensions for PDF files, XML files, and CMS signed data for signing using advanced electronic signatures (as explained in Chapter \ref{eidas}), are PAdES, XAdES, and CAdES, respectively.
They are defined in the respective specifications by the ETSI \footnote{Specification for PAdES is available at \url{https://www.etsi.org/deliver/etsi_ts/102700_102799/10277801/01.01.01_60/ts_10277801v010101p.pdf}} \footnote{Specification for XAdES is available at \url{https://uri.etsi.org/01903/v1.3.2/ts_101903v010302p.pdf}} \footnote{Specification for CAdES is available at \url{https://www.etsi.org/deliver/etsi_ts/101700_101799/101733/02.01.01_60/ts_101733v020101p.pdf}}.

When using PAdES, signatures are always part of the PDF file itself. This differs from XAdES, where the signature can be either part of the XML or provided as a separate file, or CAdES where signature is always in the form of binary data and it is up to the software.

All of them support multiple signatures applied in succession. Implementation of the PAdES is rather portable since any compatible PDF file reader or editor will work with the signatures in a similar way - signature created in one should be easily verifiable in any other. XAdES and CAdES, however, are usually tied to specific implementation or standard that is built on top of them.

@@

The ability to retain signed documents for a prolonged period of time is covered in these standards under optional Long-Term Validation (LTV) that allows archiving of documents for many years, possibly decades.
If enabled, a trustworthy timestamp is required as validity is verified for the time the signature (and therefore timestamp) was created.
The expiration time is then limited by the validity of the timestamp certificate.

\section{Software engineering}

@@

TODO: OSS Sustainability

\chapter{Goals}

@@

TODO: Website - multilingual, editable by the community
TODO: Global status - world map

Lorem ipsum dolor sit amet, consectetuer adipiscing elit.
Integer lacinia, nulla porta varius tempus, lacus metus blandit
lorem, a rutrum justo wisi id sapien. Integer risus libero,
feugiat eleifend, ornare ac, volutpat nec, sem. In facilisis,
quam eu elementum aliquet, lorem quam euismod dui, aliquet
laoreet purus ipsum ac quam.

\chapter{Analysis}

\section{Review of existing software}

TODO: Explain how they were picked, list them with info on the considered trait.

TODO: Maybe add a simple comparison table? Consider: - format support, commercial, localization, eIDAS, web, required setup, advanced features

\subsection{Desktop application JSignPdf}

\subsection{Web application zep.disig.sk}

\subsection{PDF viewer Adobe Acrobat Reader DC}

\subsection{Commercial desktop application D.PDF Signer}

\section{Modularity}

>>>>

\documentclass[thesismargins, english, thesislinespacing, onelinechapterstyle, upjsfrontpage]{rnthesis}
\usepackage[english]{babel}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}

@@

\usepackage{xparse}
% Rotation: \rot[<angle>][<width>]{<stuff>}
\NewDocumentCommand{\rot}{O{60} O{1em} m}{\makebox[#2][l]{\rotatebox{#1}{#3}}}%

\title{Replacing handwritten signatures with open electronic signature software}
\author{Jakub Ďuraš}
\iffalse % Not part of the version for ŠVK
\typprace{bachelor's}
\fi
\inytypprace{Student Scientific Conference}
\rok{2020}
\odbor{Applied Informatics}
\miesto{Košice}
\veduci{RNDr. Viliam Kačala}
\pracovisko{Institute of Computer Science}
\iffalse % Not part of the version for ŠVK
\podakovanie{
  I would like to thank my supervisor Viliam Kačala for his continuous feedback and patience with my question.

  I would also like to acknowledge help received from Pavol Sokol and Laura Rózenfeldová on legal questions, and access to various resources from Gabriel Semanišin.

  Finally, I would like to thank my family and friends who contributed in various ways to my thesis.
}
\fi

\iffalse % Not part of the version for ŠVK
\pdfzadanie{zadanie.pdf}
\fi

\abstract{
With the recent changes in the legal status of electronic signatures in the world, there is an opportunity to expand the use of electronic signatures as an alternative to handwritten signatures outside the usual areas.
Electronic signatures are becoming increasingly important in the age of electronic communication when handwritten signatures are unpractical due to the need to print, personally exchange, or send signed documents.
We devote our attention to the exploration of legal preconditions relevant for the EU, technological preconditions, review of the existing software, and proposal of our own software platform.
The result is open, user-friendly software for Windows, Mac OS X, and Linux.
Using the application, it is possible, following the eIDAS Regulation, to sign and verify the signature of any document with the legal effect of an attested signature.
Our solution supports ID cards of the Slovak Republic with the eID chip without any modifications or settings, which enables easy use outside the usual use in web applications of the state.
The application is not tied only to the Slovak eID but is ready for further use with various devices in different countries and as a platform for other types of signatures.
Another part is a website that, in addition to the possibility of downloading software and brief documentation, provides general information about electronic signatures so that the project is as accessible to the public as possible.
We concentrate on simple extensibility of the program via language translations, documentation, functionality, and general information.
}

@@

\abstrakt{
So zmenami v právnom statuse elektronických podpisov vo svete sa vytvára príležitosť rozšíriť elektronické podpisovanie ako náhradu pre vlastnoručné podpisy mimo obvyklých oblastí.
Elektronické podpisy sú dôležité čoraz viac v dobe elektronickej komunikácie kedy sú vlastnoručné podpisy nepraktické z dôvodu nutnosti tlače, osobného kontaktu alebo zasielania podpísaných dokumentov.
V práci sa venujeme právnym predpokladom relevantným pre EÚ, technologickým predpokladom, prehľadu v súčasnosti dostupného softvéru a návrhu vlastnej softvérovej platformy.
Výsledkom je otvorený, používateľsky prívetivý softvér pre platformy Windows, Mac OS X a Linux.
Pomocou aplikácie je možné, v súlade s nariadením eIDAS, podpísať a overiť podpis ľubovoľného dokumentu, ktorý má rovnaký právny účinok ako osvedčený podpis.
Naše riešenie podporuje občianske preukazy Slovenskej Republiky s čipom eID bez akýchkoľvek úprav alebo nastavení, čím sa umožní ich jednoduché využitie mimo obvyklého použitia vo webových aplikáciách štátu.
Aplikácia nie je viazaná iba na slovenské eID, ale je pripravená na ďalšie použitie s rôznymi zariadeniami v rôznych štátoch a ako platforma pre iné druhy podpisov.
Súčasťou je aj webová stránka, ktorá okrem možnosti stiahnutia softvéru a stručnej dokumentácie poskytuje všeobecné informácie o elektronických podpisoch aby bol projekt čo najviac prístupný verejnosti.
Dôraz kladieme na jednoduchú rozšíriteľnosť programu vo forme jazykových prekladov, dokumentácie, funkcionality a všeobecných informácií.
}

@@

\begin{document}
\maketitle

\newpage

\tableofcontents
\iffalse % Not part of the version for ŠVK
\listoffigures
\listoftables
\fi

@@

Electronic signatures have been around for a while.
Their practical usability, however, is influenced by their legal status and available infrastructure.
With the recent changes in the relevant law around the world and the European Union specifically, they see increased adoption.

While, specifically in the European Union, they seem to be relatively extensively used in communication with the government, they do not seem to be widely adopted in an exchange between individuals or smaller organizations.
The problem is handwritten signatures are often unpractical for electronic communication due to the need to print, personally exchange, or send signed documents.

Our goal is to explore an accessible way to use electronic signatures.
In the beginning, we go over the preconditions that make it a viable alternative to the handwritten signatures.
We cover the legal preconditions relevant for the European Union and Slovakia in particular.
Also, we explain technical preconditions focusing on the cryptographic concepts like asymmetric cryptography, timestamping, hashing, and standards that comply with the law within the European Union.
Secondly, we do a review of the existing software.
We go quickly over several applications that are meant for basic use.
We compare them using several criteria to find out the strengths and weaknesses of the current implementations and identify a potential space for improvements.
Finally, we devote our attention to our own implementation.
We explain the proposed desktop software and go into details on software engineering aspects.
Specifically, modularity, architecture, usability, localization, testing, and automation.
In the end, we compare it with the existing software and identify possible future improvements.
Another result we focus on is a supporting website and how it helps with the use of the software.
All in a way open for extensibility by the community.

\chapter{Preconditions and theory}

There are several preconditions that allow us to consider electronic signatures as an alternative.
Firstly, the current state of the law and recent changes relating to the use of electronic signatures, secondly, well-established concepts in cryptography like \pojem{asymmetric cryptography} or \pojem{hashing}, and lastly, expertise from the field of software engineering - design, implementation, and maintenance of computer software.
We will concetrate on the first two in relation to the electronic signatures, and, where applicable, open-source desktop computer software.

\iffalse % Not part of the version for ŠVK
TODO: Think about adding section on handwritten signatures
\fi

@@

We are considering law applicable locally in the Slovak Republic (SR).
Its law is greatly influenced by the European Union (EU), being its member since 2004, and by the rest of the world.
We can assume this is at least partially applicable outside of the SR as well.

@@

Signatures are an essential part of the written legally binding documents like contracts.
They are permanently affixed to the document and are supposed to uniquely identify the person and its deliberate, informed consent.
As can be seen in the Slovak Civil Code, § 40, a written legal act is valid if signed by the acting person \cite{civilcode}.
Our law often explicitly requires signatures and further clarifies their expected use.
For example, when selling an enterprise, looking at the Commercial Code, § 476, the contract requires written form and attested signatures of the seller and the buyer \cite{commercialcode}.

Signature is considered an "attested signature"\footnote{Translation from "osvedčený podpis" as used in the Slovak law.} if it is verified by an authorized third party.
This process is referred to as legalization in Notary Law, § 56, and its purpose is to attest information that could form the basis for the exercise of rights or which could cause legal consequences \cite{notarylaw}.

With regards to electronic signatures, law within the EU used to differ, with the law applicable in the SR being now-repealed \pojem{Act No. 215/2002 Coll.}. This situation changed with the EU Regulation eIDAS.

@@

\pojem{Advanced electronic signatures} are a subset of electronic signatures that have to \textbf{uniquely link and identify} the signature author, be created in a way that is \textbf{possible only by them} and any \textbf{changes to the signed document have to be detectable} \cite{eidas}.
From the technical point of view, standards \pojem{PAdES}, \pojem{XAdES}, and \pojem{CAdES} specified by the European Telecommunications Standards Institute (ETSI) comply with these requirements. We explain why this is the case in Section \ref{cryptography}.

\pojem{Qualified electronic signatures} (QES) are a subset of \pojem{advanced electronic signatures} with more specific requirements that can be found in annex 1 of the eIDAS regulation\footnote{Available at https://eur-lex.europa.eu/legal-content/EN/TXT/HTML/?uri=CELEX:32014R0910\#d1e32-111-1.}.
In more practical terms, such \pojem{advanced electronic signatures} are created with a qualified device using a qualified certificate and a qualified trust service.
Qualified, in this case, means authorized for such use by the legal authorities.
These devices, \pojem{certificates}, and trust services are explicitly listed by the EU. In many cases indirectly by listing the organization (and its items) from a specific country.

Furthermore, a qualified electronic signature has the legal effect of a handwritten signature \cite{eidas} and attested handwritten signatures in the SR.

Mentioned terminology is used when considering a natural person. Legal entities are able to use \pojem{electronic seals}, which are, from the technical point of view and for our purposes, identical.

Overall, electronic signatures are supposed to be, within the EU, interoperable and transparent alternative to handwritten signatures.

\subsection{Software copyright} \label{copyright}

@@

Such right can be exercised to license the software under either \pojem{proprietary} or \pojem{open-source} license.
\pojem{Proprietary} meaning under the exclusive legal right of the author, typically also \textbf{confidential and distributed as a paid product}.
\pojem{Open-source} meaning having its source code \textbf{freely available}, typically also \textbf{distributed for free} and without any liability (the software is provided "as is").

Motivation to license the software under an \pojem{open-source} license can differ, and so do such licenses.
While a \pojem{proprietary} license is usually made specifically for that entity and its interests, \pojem{open-source} licenses tend to be reused between different authors.
This means that consumers of the software can quickly recognize their rights and responsibilities if they decide to use, modify, or distribute the software.
In general, we can classify the \pojem{open-source} licenses into two categories: \pojem{permissive} and \pojem{copyleft}.

\pojem{Copyleft} license, in general, requires the user to publish the modified work under the compatible (free) license.
As a result, it forces them to extend the rights they have received onto others and, in turn, somehow hinder usability in software licensed under a different license.
A popular example of such license is the GNU General Public License(GPL) and its derivatives like GNU Lesser GPL (LGPL) or GNU Affero GPL (AGPL).

\pojem{Permissive} licenses, on the other hand, do not have such a requirement in place and are more suitable for potential commercialization.
They typically allow commercial use, modifications, distribution or sublicensing as long as the author is not held liable, and the original copyright and license are distributed with the software.
Often used permissive licenses are MIT License, Apache License, or BSD License.

It is not uncommon to see \pojem{hybrid licensing} - licensing under more than one license, with one being typically \pojem{open-source} and one \pojem{proprietary}.
In this scenario, users can choose which license they want to use based on their needs.

@@

As we have already mentioned, software licenses can influence whether and how it can be used as a part of different software.
Both when we are the ones using someone else's software and when someone else is using our software.
Generally speaking, the more open the license is, the higher the probability it will not prevent adoption for legal reasons.
Various possible combinations can be seen in Figure \ref{fig:license-compatibility}, where greyed out options mean such combination is not possible unless a separate licensing agreement is reached with the copyright owner.

\obrazok{./figures/license-compatibility}{Schematic representation of license directionality. \osoba{Morin 2012} \cite{licensing}}{license-compatibility}

@@

Previously mentioned standards \pojem{PAdES}, \pojem{XAdES}, and \pojem{CAdES} define how \pojem{asymmetric cryptography}, \pojem{hashing}, or \pojem{timestamping} should be used to comply with the standards.
We go over a these topics and why they are able to satisfy the requirements for QES mentioned in the Section \ref{eidas}.

@@

There are several issues that make symmetric encryption unsuitable.
Let us imagine we are creating a key for each pair of users in our network - imaginary country Isle of Alices and Bobs - for symmetric encryption and decryption of documents shared between each Alice and Bob.

\textbf{Secure distribution of keys} seems to be feasible when we are adding a new user to our network.
We require each user to personally come and pick up a physical device that contains pre-generated keys.
Changes made after that, however, would mean we either need to require a periodic personal visit to update the saved keys or transmission of these keys over some medium, which can be potentially unsafe.
Any new members of the network are unable to participate until other Alices and Bobs get their updates.

@@

which means that even our little imaginary country with a population of only 70 000 requires roughly 2.5 billion keys. However, since our signature system needs to be portable, we will need to count also with a population of other countries, and so the number of keys grows very quickly.

\iffalse % Not part of the version for ŠVK
\begin{figure}[!htb]
  \centering

@@

  \centerline{\parbox{13cm}{\sl \caption{Cubic growth of the number of keys.}}}
\end{figure}
\fi

@@

These issues can be addressed with \pojem{asymmetric cryptography}.
In general, we assign \textbf{a pair of keys} to each person, one of them used for encryption, known as a \pojem{public key}, another used for decryption, known as a \pojem{private key}.
One party can encrypt, and the other is the only one that can decrypt.

\pojem{Digital signatures} are specific \pojem{asymmetric cryptography} algorithms.
The signer is the only one who possesses the \pojem{private key} used for signing, and others can use the \pojem{public key} for signature verification.
We \textbf{sign} the document, including the other parties' signature.

\iffalse % Not part of the version for ŠVK
\obrazok{./figures/digital-signature-scheme}{Principle of digital signatures which involves signing and verifying a message (Paar 2009). TODO: I don't have rights to this image!!!}{pubkey-scheme}
\fi

Practical, widely used implementation of this idea is the \pojem{RSA} signature scheme that relies on the integer factorization problem (difficulty of factoring a product of two large primes) published in 1978 \cite{rsa}.

In Slovakia, the private key is usually handed over to the user personally, saved on the ID card. This ID card (used together with a card reader) provides an API for safe access, management, and cryptoprocessing.
Such a device is in general called a \pojem{Hardware Security Module} (HSM), and the API relevant for this thesis is \pojem{PKCS \#11} \footnote{PKCS \#11 specification available at \url{http://docs.oasis-open.org/pkcs11/pkcs11-base/v2.40/os/pkcs11-base-v2.40-os.html}.}.

We are left with the \textbf{problem of public key distribution} - even though \pojem{public-key schemes} do not require a \textit{secure channel}, they require \textit{authenticated channels} for the distribution of the public keys \cite{cryptotxtbook}.

A common solution is the use of \pojem{certificates}. \pojem{Certificates} are essentially \pojem{digital signatures} with metadata that can be used to establish the validity of the received \pojem{public key} before it is used.
Distribution of such \pojem{certificates} is the responsibility of a \pojem{Certification Authority} (CA) - a third party that all users in the network trust.
In the case of the Slovak ID cards, it is Disig - SVK eID Accredited CA, issued by the National Security Authority\footnote{Translation from "Národný bezpečnostný úrad".}.
Such \pojem{certificates}, therefore, \textbf{reliably uniquely link and identify} the \pojem{public key} owners, as can be seen in Figure \ref{fig:certificate-scheme}\footnote{Created by an unknown author, licensed under the CC BY-SA 3.0 license.}.
Commonly used standard for public key certificates is X.509\footnote{Specification available under RFC5280 at \url{https://tools.ietf.org/html/rfc5280}.}.

\obrazok{./figures/certificate-scheme}{Application of digital signature with certificate.}{certificate-scheme}

\iffalse % Not part of the version for ŠVK
TODO: Mention EU LOTL
\fi

@@

It is often useful to map data of arbitrary size to a fixed-length (typically short) set of bits.
This process is commonly called \pojem{hashing}, a function that can be used for it \pojem{hash function}, and output of such function \pojem{hash}.

In cryptography, we are considering only \pojem{hash functions} that can also satisfy these additional \textbf{requirements}:

\begin{enumerate}
  \item \textbf{Deterministic} - same input data are always mapped to the same \pojem{hash}.
  \item \textbf{One-way} - it should be practically impossible to retrieve input data for a given \pojem{hash}.
  \item \textbf{Resistant to collisions} - it should be practically impossible to find two inputs with the same \pojem{hash}.
  \item Even small change to the input should lead to significantly different \pojem{hash} - so-called \textbf{avalanche effect}.
\end{enumerate}

In the context of \pojem{digital documents}, we can use such \pojem{hash function} to generate a \pojem{hash} for the document.
This \pojem{hash} can represent a document of any size and is used during the creation of a \pojem{digital signature} instead of the document itself.
Any further changes to the document will, therefore, lead to a significantly different \pojem{hash}, which means the \pojem{digital signature} will no longer be valid, and these \textbf{changes are detected}.
For this purpose, only \pojem{hash functions} that are reasonably fast can be used, so that we can utilize them on documents of variable size.

\iffalse % Not part of the version for ŠVK
TODO: Add scheme of document > hash - valid signature, change in document > different hash - invalid signature.
\fi

An example of commonly used \pojem{hash function} family is \pojem{Secure Hash Algorithms} (SHA), with \pojem{hash functions} relevant to this thesis being mainly SHA-2\footnote{SHA-2 specification available in FIPS PUB 180-4 at \url{https://nvlpubs.nist.gov/nistpubs/FIPS/NIST.FIPS.180-4.pdf}.}.

@@

A common part of the paper signed documents is the date when they were signed.
Their digital counterpart is PKI-based \pojem{trusted timestamping} - a process of \pojem{digitally signing} date and time of document creation or modification.
No one should be able to tamper with such \pojem{timestamp} - not even the author of the document.
Additionally, we should not be able to choose arbitrary date and time ourselves, but it should represent the actual time at the signing.

To achieve this, a trusted third party - \pojem{Timestamping Authority} (TSA) - is the one that will sign the \pojem{hash} of the document together with the current date and time.
This process can be seen in Figure \ref{fig:trusted-timestamping}\footnote{Schema created by Bart Van den Bosch, licensed under the CC BY-SA 2.0 BE license.}.
Verifying can be done by decrypting this information using the \pojem{public key} and comparing the \pojem{hashes}. Any attempt at \textbf{tampering is therefore detected}.

@@

The commonly used standard for the \pojem{Public Key Infrastructure} (PKI) based timestamping is part of the X.509\footnote{Specification available under RFC3161 at \url{https://www.ietf.org/rfc/rfc3161.txt}.}.

@@

Extensions for PDF files, XML files, and CMS signed data for signing using \pojem{advanced electronic signatures} (as explained in Section \ref{eidas}), are \pojem{PAdES}, \pojem{XAdES}, and \pojem{CAdES}, respectively.
They are defined in the respective specifications by the ETSI in \cite{pades}, \cite{xades}, and \cite{cades}.

When using PAdES, signatures are always part of the PDF file itself.
This differs from XAdES, where the signature can be either part of the XML or provided as a separate file, or CAdES where the signature is always in the form of binary data, and it is up to the software.

All of them support multiple signatures applied in succession.
\iffalse % Not part of the version for ŠVK
Implementation of the PAdES is rather portable since any compatible PDF file reader or editor will work with the signatures in a similar way - signature created in one should be easily verifiable in any other. XAdES and CAdES, however, are usually tied to specific implementation or standard that is built on top of them.
\fi

@@

The ability to keep signed documents valid for a prolonged period of time is covered in these standards under optional \pojem{Long-Term Validation} (LTV) that allows \textbf{archiving} of documents for many years, possibly decades.
If enabled, a trustworthy timestamp is required as validity is verified for the time the signature (and therefore timestamp) was created.
The expiration time is then limited by the validity of the timestamp certificate.

\iffalse % Not part of the version for ŠVK
\section{Software engineering}

We will consider Ian Sommerville's definition of software engineeing - "an engineering discipline that is concerned with all aspects of software production from initial conception to operation and maintenance" \cite{sommerville}.
Aspects we will focus on are design, implementation, testing and releasing, and maintenance.
Small

@@

TODO: OSS Sustainability
\fi

\iffalse % Not part of the version for ŠVK
\chapter{Goals}

@@

TODO: Website - multilingual, editable by the community
TODO: Global status - world map
\fi

\chapter{Review of existing software} \label{review}

We focus only on simple applications made mostly for the signing of the documents.
During our review, we have used several PDF files that we have tried to sign using QES available on the SR eID.
We have tried to include software of all types in our review, preferring the most popular, locally available ones, since those should be prepared for the use with the SR eID.
Excluded are all complex applications with advanced features and high price tag.
We have picked one open-source desktop application, one web application - Software as a Service (SaaS), one PDF viewer, and one simple paid desktop application. Criteria we focus on are:
